---
title: "Data preprocessing"
author: "Ethan Naegele"
date: "2024-05-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(gam)
library(lubridate)
library(caret)
library(splines)
library(locfit)
library(locpol)
library(BradleyTerry2)
library(mgcv)
library(e1071)
```

NOTE: NOT USING BPS PER MATCH SINCE THAT HAS 1 LESS OBSERVATION THAN BPS PER SET

```{r}
data_2019 <- read_csv('atp_matches_2019.csv')
data_2020 <- read_csv('atp_matches_2020.csv')
data_2021 <- read_csv('atp_matches_2021.csv')
data_2022 <- read_csv('atp_matches_2022.csv')
data_2023 <- read_csv('atp_matches_2023.csv')
```


```{r}
# Make sure to copy this to another file and change your path when you run this code on your own computer!

first_serve_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/1st_serve_won_pct_all_2019.csv')

second_serve_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/2nd_serve_won_pct_all_2019.csv')

aces_per_set_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/aces_per_set_all_2019.csv')

bps_per_match_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/bps_per_match_all_2019.csv')

bps_per_set_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/bps_per_set_all_2019.csv')

dfs_per_set_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/dfs_per_set_all_2019.csv')

pts_won_per_rtn_game_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/pts_won_per_rtn_game_all_2019.csv')

rtn_in_play_pts_won_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/rtn_in_play_pts_won_all_2019.csv')

tot_2nd_srv_in_pl_pts_w_all_2019 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/tot_2nd_srv_in_pl_pts_w_all_2019.csv')


```




```{r}
vectors_list <- list(first_serve_all_2019$name, 
                     second_serve_all_2019$name, 
                     aces_per_set_all_2019$name,
                     bps_per_match_all_2019$name, 
                     bps_per_set_all_2019$name,
                     dfs_per_set_all_2019$name, 
                     pts_won_per_rtn_game_all_2019$name,
                     rtn_in_play_pts_won_all_2019$name, 
                     tot_2nd_srv_in_pl_pts_w_all_2019$name)

# Find the intersection of all vectors
common_names <- Reduce(intersect, vectors_list)
common_names
```



```{r}
left_join_by_name <- function(df1, df2) {
  left_join(df1, df2, by = "name")
}

rename_and_select <- function(df, new_value_name) {
  df %>%
    select(name, value) %>%
    rename(!!new_value_name := value)
}
```

```{r}

data_frames_list <- list(first_serve_all_2019 = first_serve_all_2019, 
                         second_serve_all_2019 = second_serve_all_2019, 
                         aces_per_set_all_2019 = aces_per_set_all_2019,
                         bps_per_set_all_2019 = bps_per_set_all_2019,
                         dfs_per_set_all_2019 = dfs_per_set_all_2019, 
                         pts_won_per_rtn_game_all_2019 = pts_won_per_rtn_game_all_2019,
                         rtn_in_play_pts_won_all_2019 = rtn_in_play_pts_won_all_2019, 
                         tot_2nd_srv_in_pl_pts_w_all_2019 = tot_2nd_srv_in_pl_pts_w_all_2019)


# Apply the function to each data frame and store the results in a list
renamed_data_frames <- mapply(rename_and_select, data_frames_list, names(data_frames_list), SIMPLIFY = FALSE)

# Perform left joins iteratively
merged_data_2019 <- Reduce(left_join_by_name, renamed_data_frames)
```



```{r}
covs_2019 <- na.omit(merged_data_2019)
```

```{r}
# Make sure to copy this to another file and change your path when you run this code on your own computer!

first_serve_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/1st_serve_won_pct_all_2020.csv')

second_serve_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/2nd_serve_won_pct_all_2020.csv')

aces_per_set_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/aces_per_set_all_2020.csv')

bps_per_set_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/bps_per_set_all_2020.csv')

dfs_per_set_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/dfs_per_set_all_2020.csv')

pts_won_per_rtn_game_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/pts_won_per_rtn_game_all_2020.csv')

rtn_in_play_pts_won_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/rtn_in_play_pts_won_all_2020.csv')

tot_2nd_srv_in_pl_pts_w_all_2020 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/tot_2nd_srv_in_pl_pts_w_all_2020.csv')


```


```{r}

data_frames_list <- list(first_serve_all_2020 = first_serve_all_2020, 
                         second_serve_all_2020 = second_serve_all_2020, 
                         aces_per_set_all_2020 = aces_per_set_all_2020,
                         bps_per_set_all_2020 = bps_per_set_all_2020,
                         dfs_per_set_all_2020 = dfs_per_set_all_2020, 
                         pts_won_per_rtn_game_all_2020 = pts_won_per_rtn_game_all_2020,
                         rtn_in_play_pts_won_all_2020 = rtn_in_play_pts_won_all_2020, 
                         tot_2nd_srv_in_pl_pts_w_all_2020 = tot_2nd_srv_in_pl_pts_w_all_2020)



# Apply the function to each data frame and store the results in a list
renamed_data_frames <- mapply(rename_and_select, data_frames_list, names(data_frames_list), SIMPLIFY = FALSE)

# Perform left joins iteratively
merged_data_2020 <- Reduce(left_join_by_name, renamed_data_frames)

covs_2020 <- na.omit(merged_data_2020)

```


```{r}
first_serve_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/1st_serve_won_pct_all_2021.csv')

second_serve_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/2nd_serve_won_pct_all_2021.csv')

aces_per_set_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/aces_per_set_all_2021.csv')

bps_per_set_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/bps_per_set_all_2021.csv')

dfs_per_set_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/dfs_per_set_all_2021.csv')

pts_won_per_rtn_game_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/pts_won_per_rtn_game_all_2021.csv')

rtn_in_play_pts_won_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/rtn_in_play_pts_won_all_2021.csv')

tot_2nd_srv_in_pl_pts_w_all_2021 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/tot_2nd_srv_in_pl_pts_w_all_2021.csv')


```


```{r}

data_frames_list <- list(first_serve_all_2021 = first_serve_all_2021, 
                         second_serve_all_2021 = second_serve_all_2021, 
                         aces_per_set_all_2021 = aces_per_set_all_2021,
                         bps_per_set_all_2021 = bps_per_set_all_2021,
                         dfs_per_set_all_2021 = dfs_per_set_all_2021, 
                         pts_won_per_rtn_game_all_2021 = pts_won_per_rtn_game_all_2021,
                         rtn_in_play_pts_won_all_2021 = rtn_in_play_pts_won_all_2021, 
                         tot_2nd_srv_in_pl_pts_w_all_2021 = tot_2nd_srv_in_pl_pts_w_all_2021)



# Apply the function to each data frame and store the results in a list
renamed_data_frames <- mapply(rename_and_select, data_frames_list, names(data_frames_list), SIMPLIFY = FALSE)

# Perform left joins iteratively
merged_data_2021 <- Reduce(left_join_by_name, renamed_data_frames)

covs_2021 <- na.omit(merged_data_2021)

```


```{r}
first_serve_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/1st_serve_won_pct_all_2022.csv')

second_serve_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/2nd_serve_won_pct_all_2022.csv')

aces_per_set_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/aces_per_set_all_2022.csv')

bps_per_set_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/bps_per_set_all_2022.csv')

dfs_per_set_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/dfs_per_set_all_2022.csv')

pts_won_per_rtn_game_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/pts_won_per_rtn_game_all_2022.csv')

rtn_in_play_pts_won_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/rtn_in_play_pts_won_all_2022.csv')

tot_2nd_srv_in_pl_pts_w_all_2022 <- read_csv('~/Documents/R files/Stat_274_project/covariates_data/tot_2nd_srv_in_pl_pts_w_all_2022.csv')


```

```{r}

data_frames_list <- list(first_serve_all_2022 = first_serve_all_2022, 
                         second_serve_all_2022 = second_serve_all_2022, 
                         aces_per_set_all_2022 = aces_per_set_all_2022,
                         bps_per_set_all_2022 = bps_per_set_all_2022,
                         dfs_per_set_all_2022 = dfs_per_set_all_2022, 
                         pts_won_per_rtn_game_all_2022 = pts_won_per_rtn_game_all_2022,
                         rtn_in_play_pts_won_all_2022 = rtn_in_play_pts_won_all_2022, 
                         tot_2nd_srv_in_pl_pts_w_all_2022 = tot_2nd_srv_in_pl_pts_w_all_2022)



# Apply the function to each data frame and store the results in a list
renamed_data_frames <- mapply(rename_and_select, data_frames_list, names(data_frames_list), SIMPLIFY = FALSE)

# Perform left joins iteratively
merged_data_2022 <- Reduce(left_join_by_name, renamed_data_frames)

covs_2022 <- na.omit(merged_data_2022)

```


```{r}
# reduced dataframes
data_2019r <- data_2019 %>% 
  select(tourney_date, winner_name,
        loser_name, winner_rank, loser_rank,
         w_ace, w_df, w_1stIn, w_1stWon, w_svpt, w_2ndWon, w_bpFaced, w_bpSaved,
         l_ace, l_df, l_1stIn, l_1stWon, l_svpt, l_2ndWon, l_bpFaced, l_bpSaved
         ) %>% 
  na.omit()

data_2020r <- data_2020 %>% 
  select(tourney_date, winner_name,
        loser_name, winner_rank, loser_rank,
         w_ace, w_df, w_1stIn, w_1stWon, w_svpt, w_2ndWon, w_bpFaced, w_bpSaved,
         l_ace, l_df, l_1stIn, l_1stWon, l_svpt, l_2ndWon, l_bpFaced, l_bpSaved
         ) %>% 
  na.omit()

data_2021r <- data_2021 %>% 
  select(tourney_date, winner_name,
        loser_name, winner_rank, loser_rank,
         w_ace, w_df, w_1stIn, w_1stWon, w_svpt, w_2ndWon, w_bpFaced, w_bpSaved,
         l_ace, l_df, l_1stIn, l_1stWon, l_svpt, l_2ndWon, l_bpFaced, l_bpSaved
         ) %>% 
  na.omit()

data_2022r <- data_2022 %>% 
  select(tourney_date, winner_name,
        loser_name, winner_rank, loser_rank,
         w_ace, w_df, w_1stIn, w_1stWon, w_svpt, w_2ndWon, w_bpFaced, w_bpSaved,
         l_ace, l_df, l_1stIn, l_1stWon, l_svpt, l_2ndWon, l_bpFaced, l_bpSaved
         ) %>% 
  na.omit()

data_2023r <- data_2023 %>% 
  select(tourney_date, winner_name,
        loser_name, winner_rank, loser_rank,
         w_ace, w_df, w_1stIn, w_1stWon, w_svpt, w_2ndWon, w_bpFaced, w_bpSaved,
         l_ace, l_df, l_1stIn, l_1stWon, l_svpt, l_2ndWon, l_bpFaced, l_bpSaved
         ) %>% 
  na.omit()


```



```{r}
# now, consider only the matches for which we have covariate data available for both players

names_vec <- intersect(c(data_2020r$loser_name, data_2020r$winner_name), covs_2019$name)
data_2020_train <- data_2020r %>%
  filter(winner_name %in% names_vec & loser_name %in% names_vec)
  
names_vec <- intersect(c(data_2021r$loser_name, data_2021r$winner_name), covs_2020$name)
data_2021_train <- data_2021r %>%
  filter(winner_name %in% names_vec & loser_name %in% names_vec)

names_vec <- intersect(c(data_2022r$loser_name, data_2022r$winner_name), covs_2021$name)
data_2022_train <- data_2022r %>%
  filter(winner_name %in% names_vec & loser_name %in% names_vec)
  
  

```



```{r}
# There are some columns in the dataframes that are character-valued that contain percentages. We convert those to decimal proportions.
convert_percentage <- function(df, columns) {
  df %>%
    mutate(across(all_of(columns), ~ as.numeric(sub("%", "", .)) / 100))
}

# List of years to process
years <- 2019:2022

# Create a named list of data frames
covs_list <- list(covs_2019 = covs_2019, covs_2020 = covs_2020, covs_2021 = covs_2021, covs_2022 = covs_2022)

# Process each year
for (year in years) {
  # Generate column names for the current year
  columns_to_convert <- c(paste0("first_serve_all_", year),
                          paste0("second_serve_all_", year),
                          paste0("rtn_in_play_pts_won_all_", year),
                          paste0("tot_2nd_srv_in_pl_pts_w_all_", year))
  
  # Convert percentages in the corresponding data frame
  covs_list[[paste0("covs_", year)]] <- convert_percentage(covs_list[[paste0("covs_", year)]], columns_to_convert)
}

# Assign back the processed data frames to the original names
covs_2019 <- covs_list$covs_2019
covs_2020 <- covs_list$covs_2020
covs_2021 <- covs_list$covs_2021
covs_2022 <- covs_list$covs_2022
```



```{r}
# combine the dataframes
data_train <- rbind(data_2020_train, data_2021_train, data_2022_train)
```


```{r}
# the dates are not formatted well in the original dataframe, reformatting using lubridate
data_train$tourney_date <- ymd(data_train$tourney_date)

# Ensure the dates are in Date class (in case they need formatting to mdy, they remain Date class)
data_train$tourney_date <- as.Date(data_train$tourney_date, format = "%Y-%m-%d")
```

```{r}
# create a variable keeping track of the year that the match took place in

data_train$year <- year(data_train$tourney_date)
```

Determining player 1 and player 2, which determines how we code the outcome (0 or 1)

```{r}
data_train$player1_name <- ifelse(data_train$winner_rank < data_train$loser_rank, data_train$winner_name, data_train$loser_name) # denote player 1 by whoever has higher rank, for sake of consistency

data_train$player2_name <- ifelse(data_train$winner_rank < data_train$loser_rank, data_train$loser_name, data_train$winner_name)

 # Create a binary outcome variable (1 if player1 wins, 0 if player2 wins)
data_train$outcome <- ifelse(data_train$winner_rank < data_train$loser_rank, 1, 0)


data_train$player1_name <- as.factor(data_train$player1_name)
data_train$player2_name <- as.factor(data_train$player2_name)

```

I want to get the covariates we had before in the dataframe as well


```{r}

# Function to create player stats for a given year
create_player_stats <- function(year) {
  data <- get(paste0("data_", year, "r"))
  
  data_winner <- data %>%
    select(winner_name, w_ace, w_df, w_1stIn, w_1stWon, w_svpt, w_2ndWon, w_bpFaced, w_bpSaved) %>%
    rename(player_name = winner_name, player_ace = w_ace, player_df = w_df, 
           player_1stIn = w_1stIn,
           player_1stWon = w_1stWon,
           player_svpt = w_svpt,
           player_2ndWon = w_2ndWon,
           player_bpFaced = w_bpFaced,
           player_bpSaved = w_bpSaved
           ) %>% 
    mutate(player_1stEfficiency = player_1stWon / pmax(player_1stIn, 1)) %>% 
    mutate(player_2ndEfficiency = player_2ndWon / pmax(player_svpt - player_1stIn, 1))
  
  data_loser <- data %>%
    select(loser_name, l_ace, l_df, l_1stIn, l_1stWon, l_svpt, l_2ndWon, l_bpFaced, l_bpSaved) %>%
    rename(player_name = loser_name, player_ace = l_ace, player_df = l_df,
           player_1stIn = l_1stIn,
           player_1stWon = l_1stWon,
           player_svpt = l_svpt,
           player_2ndWon = l_2ndWon,
           player_bpFaced = l_bpFaced,
           player_bpSaved = l_bpSaved
           ) %>% 
    mutate(player_1stEfficiency = player_1stWon / pmax(player_1stIn, 1)) %>% 
    mutate(player_2ndEfficiency = player_2ndWon / pmax(player_svpt - player_1stIn, 1))
  
  # Combine both datasets
  combined_data <- bind_rows(data_winner, data_loser)
  
  # Aggregate the data
  player_stats <- combined_data %>%
    group_by(player_name) %>%
    summarise(
      total_aces = sum(player_ace, na.rm = TRUE),
      total_dfs = sum(player_df, na.rm = TRUE),
      average_2ndWon = mean(player_2ndWon, na.rm = TRUE),
      average_1stEfficiency = mean(player_1stEfficiency, na.rm = TRUE),
      average_2ndEfficiency = mean(player_2ndEfficiency, na.rm = TRUE),
      total_bpSaved = sum(player_bpSaved, na.rm = TRUE),
      total_bpFaced = sum(player_bpFaced, na.rm = TRUE),
      .groups = 'drop'  # This drops the grouping structure after summarising
    )
  
  # Calculate ace to double fault ratio
  player_stats <- player_stats %>%
    mutate(ace_df_ratio = total_aces / pmax(total_dfs, 1)) %>% 
    mutate(bp_saved_faced_ratio = total_bpSaved / pmax(total_bpFaced, 1))
  
  covs <- get(paste0("covs_", year))
  player_stats <- player_stats %>% 
    filter(player_name %in% covs$name)
  
  assign(paste0("player_stats_", year), player_stats, envir = .GlobalEnv)
}



```


```{r}
merge_player_stats_with_covs <- function(year) {
  player_stats <- get(paste0("player_stats_", year))
  covs <- get(paste0("covs_", year))
  
  covs_merged <- covs %>%
    left_join(player_stats, by = c('name' = "player_name")) %>% 
    na.omit()
  
  
  assign(paste0("covs_", year, "_merged"), covs_merged, envir = .GlobalEnv)
}

# Create player stats and merge with covs for each year from 2019 to 2022
for (year in 2019:2022) {
  create_player_stats(year)
  merge_player_stats_with_covs(year)
}
```



I now want to add columns to data_train in the following way. If the year is Y, I want to find the winner's and loser's values for each of the columns in covs_Y-1, so for example if the year is 2020, I want to find the values of first_serve_all_2020, second_serve_all_2020, aces_per_set_all_2020, bps_per_set_all_2020, dfs_per_set_all_2020, pts_won_per_rtn_game_all_2020, rtn_in_play_pts_won_all_2020, tot_2nd_srv_in_pl_pts_w_all_2020

for both the winner and loser. 


```{r}
# SEEMS TO BE CORRECT WORKING VERSION

# Assuming data_train, covs_2019, covs_2020, and covs_2021 dataframes are already loaded

# List of covs dataframes
covs_list <- list(`2019` = covs_2019_merged, `2020` = covs_2020_merged, `2021` = covs_2021_merged)

# Function to rename columns for player1 and player2
rename_columns <- function(covs, year) {
  colnames(covs)[2:ncol(covs)] <- gsub(paste0("_all_", year), "", colnames(covs)[2:ncol(covs)])
  covs_player1 <- covs
  covs_player2 <- covs
  colnames(covs_player1)[2:ncol(covs_player1)] <- paste0("player1_", colnames(covs_player1)[2:ncol(covs_player1)])
  colnames(covs_player2)[2:ncol(covs_player2)] <- paste0("player2_", colnames(covs_player2)[2:ncol(covs_player2)])
  list(player1 = covs_player1, player2 = covs_player2)
}

# Ensure the 'name' column in each covariate dataframe is unique
covs_list <- lapply(covs_list, function(covs) {
  covs <- covs[!duplicated(covs$name), ]
  return(covs)
})

# Create an empty list to store the processed data for each year
data_train_list <- list()

# Iterate over each year and merge the covariate data
for (year in c(2020, 2021, 2022)) {
  # Filter data_train for the current year
  data_train_year <- data_train %>% filter(year == !!year)
  
  # Select the appropriate covariate dataframe
  covs <- covs_list[[as.character(year - 1)]]
  
  # Rename columns for player1 and player2
  renamed_covs <- rename_columns(covs, year - 1)
  
  # Debugging print statements
  print(paste("Processing year:", year))
  print(paste("Rows in data_train_year before merge:", nrow(data_train_year)))
  
  # Merge the player1 data
  data_train_year <- data_train_year %>%
    left_join(renamed_covs$player1, by = c("player1_name" = "name"))
  
  # Debugging print statements
  print(paste("Rows in data_train_year after player1 merge:", nrow(data_train_year)))
  
  # Merge the player2 data
  data_train_year <- data_train_year %>%
    left_join(renamed_covs$player2, by = c("player2_name" = "name"))
  
  # Debugging print statements
  print(paste("Rows in data_train_year after player2 merge:", nrow(data_train_year)))
  
  # Append the processed data for the current year to the list
  data_train_list[[as.character(year)]] <- data_train_year
}

# Combine the processed data for all years into a single dataframe
data_train_processed <- bind_rows(data_train_list)

# Display the resulting dataframe
print(head(data_train_processed))
print(paste("Rows in data_train_processed:", nrow(data_train_processed)))
print(paste("Rows in original data_train:", nrow(data_train)))


```



```{r}
# we check that there should not be any additional NAs introduced
data_train_processed <-  na.omit(data_train_processed) 
```


```{r}
# trim down the number of variables

data_train_covs <- data_train_processed %>% 
  select(outcome, tourney_date, player1_name, player2_name, 
         player1_ace_df_ratio, player1_average_1stEfficiency, 
         player1_average_2ndEfficiency, player1_first_serve, 
         player1_second_serve,
         player1_aces_per_set, player1_average_2ndWon, 
         player1_bp_saved_faced_ratio, player1_bps_per_set,
         player1_dfs_per_set, player1_pts_won_per_rtn_game, 
         player1_rtn_in_play_pts_won, player1_tot_2nd_srv_in_pl_pts_w,
         
         player2_ace_df_ratio, player2_average_1stEfficiency, 
         player2_average_2ndEfficiency, player2_first_serve,
         player2_second_serve,
         player2_aces_per_set, player2_average_2ndWon, 
         player2_bp_saved_faced_ratio, player2_bps_per_set,
         player2_dfs_per_set, player2_pts_won_per_rtn_game, 
         player2_rtn_in_play_pts_won, player2_tot_2nd_srv_in_pl_pts_w
         )
```


```{r}
scale_numeric_columns <- function(df) {
  numeric_columns <- sapply(df, is.numeric)
  df[numeric_columns] <- scale(df[numeric_columns])
  return(df)
}


```


```{r}
data_train_covs <- data_train_covs %>% 
  mutate(player1_dfs_per_set_inverse = 1 / player1_dfs_per_set,
         player2_dfs_per_set_inverse = 1 / player2_dfs_per_set)
```


```{r}
# Scale numeric columns in the data_train_processed dataframe
data_train_covs <- scale_numeric_columns(data_train_covs)
```

```{r}
data_train_covs <- data_train_covs %>% 
  mutate(ace_df_ratio_diff = player1_ace_df_ratio - player2_ace_df_ratio,
         average_1stEfficiency_diff = player1_average_1stEfficiency - player2_average_1stEfficiency,
         average_2ndEfficiency_diff = player1_average_2ndEfficiency - player2_average_2ndEfficiency,
         first_serve_diff = player1_first_serve - player2_first_serve,
         second_serve_diff = player1_second_serve - player2_second_serve,
         aces_per_set_diff = player1_aces_per_set - player2_aces_per_set,
         average_2ndWon_diff = player1_average_2ndWon - player2_average_2ndWon,
         bp_saved_faced_ratio_diff = player1_bp_saved_faced_ratio - player2_bp_saved_faced_ratio,
         bps_per_set_diff = player1_bps_per_set - player2_bps_per_set,
         dfs_per_set_diff = player1_dfs_per_set - player2_dfs_per_set,
         pts_won_per_rtn_game_diff = player1_pts_won_per_rtn_game - player2_pts_won_per_rtn_game,
         rtn_in_play_pts_won_diff = player1_rtn_in_play_pts_won - player2_rtn_in_play_pts_won,
         tot_2nd_srv_in_pl_pts_w_diff = player1_tot_2nd_srv_in_pl_pts_w - player2_tot_2nd_srv_in_pl_pts_w
         )
```

```{r}
# the model with only the outcome and the covariates
data_train_model <- data_train_covs %>% 
  select(outcome, ace_df_ratio_diff, average_1stEfficiency_diff, average_2ndEfficiency_diff,
         first_serve_diff, second_serve_diff, aces_per_set_diff, average_2ndWon_diff, 
         bp_saved_faced_ratio_diff, bps_per_set_diff, dfs_per_set_diff, 
         pts_won_per_rtn_game_diff, rtn_in_play_pts_won_diff, 
         tot_2nd_srv_in_pl_pts_w_diff)
```


# Formatting the test set


```{r}
data_2023r$player1_name <- ifelse(data_2023r$winner_rank < data_2023r$loser_rank, data_2023r$winner_name, data_2023r$loser_name) # denote player 1 by whoever has higher rank, for sake of consistency

data_2023r$player2_name <- ifelse(data_2023r$winner_rank < data_2023r$loser_rank, data_2023r$loser_name, data_2023r$winner_name)

 # Create a binary outcome variable (1 if player1 wins, 0 if player2 wins)
data_2023r$outcome <- ifelse(data_2023r$winner_rank < data_2023r$loser_rank, 1, 0)


data_2023r$player1_name <- as.factor(data_2023r$player1_name)
data_2023r$player2_name <- as.factor(data_2023r$player2_name)

```

```{r}
# have to rename the levels or the classification won't work
data_2023r$outcome <- as.factor(data_2023r$outcome)
levels(data_2023r$outcome) <- c("loss", "win")
```

```{r}
names_vec <- intersect(c(data_2023r$loser_name, data_2023r$winner_name), covs_2022$name)
data_2023_test <- data_2023r %>%
  filter(winner_name %in% names_vec & loser_name %in% names_vec)
```

```{r}
# the dates are not formatted well in the original dataframe, reformatting using lubridate
data_2023_test$tourney_date <- ymd(data_2023_test$tourney_date)

# Ensure the dates are in Date class (in case they need formatting to mdy, they remain Date class)
data_2023_test$tourney_date <- as.Date(data_2023_test$tourney_date, format = "%Y-%m-%d")


# create a variable keeping track of the year that the match took place in

data_2023_test$year <- year(data_2023_test$tourney_date)
```





```{r}
# SEEMS TO BE CORRECT WORKING VERSION

# Assuming data_train, covs_2019, covs_2020, and covs_2021 dataframes are already loaded

# List of covs dataframes
covs_list <- list(`2022` = covs_2022_merged)


# Ensure the 'name' column in each covariate dataframe is unique
covs_list <- lapply(covs_list, function(covs) {
  covs <- covs[!duplicated(covs$name), ]
  return(covs)
})


# Iterate over each year and merge the covariate data
for (year in c(2023)) {
  # Filter data_train for the current year
  data_2023_test_processed <- data_2023_test %>% filter(year == !!year)
  
  # Select the appropriate covariate dataframe
  covs <- covs_list[[as.character(year - 1)]]
  
  # Rename columns for player1 and player2
  renamed_covs <- rename_columns(covs, year - 1)
  
  # Merge the player1 data
  data_2023_test_processed <- data_2023_test_processed %>%
    left_join(renamed_covs$player1, by = c("player1_name" = "name"))
  
  
  # Merge the player2 data
  data_2023_test_processed <- data_2023_test_processed %>%
    left_join(renamed_covs$player2, by = c("player2_name" = "name"))
  
}


# Display the resulting dataframe
print(head(data_2023_test_processed))
print(paste("Rows in data_train_processed:", nrow(data_2023_test_processed)))
print(paste("Rows in original data_train:", nrow(data_2023_test)))


```

```{r}
# we check that there should not be any additional NAs introduced
data_2023_test_processed <-  na.omit(data_2023_test_processed) 
```


```{r}
# trim down the number of variables

data_2023_test_covs <- data_2023_test_processed %>% 
  select(outcome, tourney_date, player1_name, player2_name, 
         player1_ace_df_ratio, player1_average_1stEfficiency, 
         player1_average_2ndEfficiency, player1_first_serve, 
         player1_second_serve,
         player1_aces_per_set, player1_average_2ndWon, 
         player1_bp_saved_faced_ratio, player1_bps_per_set,
         player1_dfs_per_set, player1_pts_won_per_rtn_game, 
         player1_rtn_in_play_pts_won, player1_tot_2nd_srv_in_pl_pts_w,
         
         player2_ace_df_ratio, player2_average_1stEfficiency, 
         player2_average_2ndEfficiency, player2_first_serve,
         player2_second_serve,
         player2_aces_per_set, player2_average_2ndWon, 
         player2_bp_saved_faced_ratio, player2_bps_per_set,
         player2_dfs_per_set, player2_pts_won_per_rtn_game, 
         player2_rtn_in_play_pts_won, player2_tot_2nd_srv_in_pl_pts_w
         )
```



```{r}
data_2023_test_covs <- data_2023_test_covs %>% 
  mutate(player1_dfs_per_set_inverse = 1 / player1_dfs_per_set,
         player2_dfs_per_set_inverse = 1 / player2_dfs_per_set)
```


```{r}
# Scale numeric columns in the data_train_processed dataframe
data_2023_test_covs <- scale_numeric_columns(data_2023_test_covs)
```

```{r}
data_2023_test_covs <- data_2023_test_covs %>% 
  mutate(ace_df_ratio_diff = player1_ace_df_ratio - player2_ace_df_ratio,
         average_1stEfficiency_diff = player1_average_1stEfficiency - player2_average_1stEfficiency,
         average_2ndEfficiency_diff = player1_average_2ndEfficiency - player2_average_2ndEfficiency,
         first_serve_diff = player1_first_serve - player2_first_serve,
         second_serve_diff = player1_second_serve - player2_second_serve,
         aces_per_set_diff = player1_aces_per_set - player2_aces_per_set,
         average_2ndWon_diff = player1_average_2ndWon - player2_average_2ndWon,
         bp_saved_faced_ratio_diff = player1_bp_saved_faced_ratio - player2_bp_saved_faced_ratio,
         bps_per_set_diff = player1_bps_per_set - player2_bps_per_set,
         dfs_per_set_diff = player1_dfs_per_set - player2_dfs_per_set,
         pts_won_per_rtn_game_diff = player1_pts_won_per_rtn_game - player2_pts_won_per_rtn_game,
         rtn_in_play_pts_won_diff = player1_rtn_in_play_pts_won - player2_rtn_in_play_pts_won,
         tot_2nd_srv_in_pl_pts_w_diff = player1_tot_2nd_srv_in_pl_pts_w - player2_tot_2nd_srv_in_pl_pts_w
         )
```

```{r}
# the model with only the outcome and the covariates
data_test_model <- data_2023_test_covs %>% 
  select(outcome, ace_df_ratio_diff, average_1stEfficiency_diff, average_2ndEfficiency_diff,
         first_serve_diff, second_serve_diff, aces_per_set_diff, average_2ndWon_diff, 
         bp_saved_faced_ratio_diff, bps_per_set_diff, dfs_per_set_diff, 
         pts_won_per_rtn_game_diff, rtn_in_play_pts_won_diff, 
         tot_2nd_srv_in_pl_pts_w_diff)
```




